ActiveCA: Time Use Data from the General Social Survey of Canada to Study Active Travel

### 2025-08-14 paper/EPB-Final-Submission.Rmd
Final version for EPB publication

---
title: "ActiveCA: Time Use Data from the General Social Survey of Canada to Study Active Travel"
author:
  - name: "Anon1"
    email: "Anon1@anon.org"
    affiliation: CODE
  - name: "Anon2"
    email: "Anon2@anon.org"
    affiliation: CODE
  - name: "Anon3"
    email: "Anon3@anon.org"
    affiliation: CODE
address:
  - code: CODE
    address: "Some school"
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: 
  "This paper describes {ActiveCA}, an open data product with Canadian time use data. {ActiveCA} is an R data package that contains analysis-ready data related to active travel spanning almost 40 years, extracted from cycles 2 (1986), 7 (1992), 12 (1998), 19 (2005), 24 (2010), 29 (2015), and 34 (2022) of the Time Use Survey (TUS) cicles from the General Social Survey (GSS). Active travel is characterized by mode, with walking being part of every cycle and bicycling starting in 1992. The attributes of active trips are the types of locations of origins and destinations, the duration of trips, and episode weights for expanding the trips to population-wide estimates. Based on the year of the survey, a variety of locations are coded. In earlier cycles, these include home, work or school, and other's home, whereas in later cycles these are augmented with locations such as grocery stores, restaurants, outdoor destinations, and others. The geographical resolution includes the province and whether the episode was in an urban or rural setting."

keywords: Active; mobility; walking; cycling; travel time; time-use; General Social Survey 
classoption:
  - Royal
  - times
bibliography: bibfile.bib
bibliographystyle: sageh
output:
  rticles::sage_article:
    keep_tex: yes
editor_options: 
  markdown: 
    wrap: 72
---

@@

{ActiveCA} provides analysis-ready data concerning active travel in
Canada spanning a period of almost 40 years, obtained from the TUS
cycles of the Canadian GSS. The GSS program is designed to provide
cross-sectional data on topics of interest to improve the well-being of
Canadians. As part of this program, every five to seven years the survey
is done on the topic of time use (TUS). Concretely, {ActiveCA} covers
Cycles 2 (1986), 7 (1992), 12 (1998), 19 (2005), 24 (2010), 29 (2015),
and 34 (2022) of the TUS. Time use data in these surveys is coded using
a very fine grain, from time spent in chores, leisure, and sleeping, to
time spent working or at school. These surveys have proved valuable in
investigations of mobility and quality of life [@spinneyTransport2009],
the relationship between active travel and transit use
[@lachapelleLonger2016], and travel behavior and time poverty
[@kimFacing2024], to name but a few examples.

Using Public Use Microdata Files (PUMFs) from the Time Use Survey (TUS),
we extracted all data necessary to characterize active travel in
Canada - specifically, episodes in which the activity involved moving
between an origin and a destination by walking or cycling. Although
Statistics Canada provides PUMFs and accompanying documentation for the
GSS program [see @statisticscanada2024], accessing and preparing these
files for analysis is not a straightforward task due to their size and
complexity. The process of extracting information of interest from the
source files is time-consuming, tedious, and challenging and/or prone to
error due to the expertise required to work with these files. To create
{ActiveCA} we selected, labelled, and processed the TUS cycles to make
them ready for analysis.

{ActiveCA} is distributed as an `R` package with a number of data objects
and their documentation. `R` packages contain code, data, and
documentation in a standardized format that can be installed by R users
via a software repository, such as CRAN (Comprehensive R Archive
Network) or GitHub, which makes them an adroit medium to distribute
analysis-ready data.

@@

The rest of this paper discusses the sources of data, and the process
implemented to retrieve and package them. Then, we explain how to use
the package and show some selected examples of analysis to whet the
imagination of potential users. This ODP provides not only data that are
easy to use, but also all the code and documentation that make this a
reproducible research project. In summary, {ActiveCA} aims to implement
and inspire the best principles of open spatial sciences
[@paez_open_2021; @brunsdon_opening_2021].

@@

Most respondents to the 2022 TUS completed it online, reflecting
Statistics Canada's effort to adapt to technological changes and growing
time demands by offering greater flexibility and convenience
[@statisticscanada2022]. While such methodological changes may affect
data comparability over time, it is not possible to determine whether
observed differences result from actual population changes or shifts in
data collection methods. Despite rigorous efforts to ensure data
quality, the use of electronic questionnaires may have influenced
estimates. Statistics Canada assessed the impact of collection mode on a
limited set of key questions, constrained by sample size. Importantly,
none of the variables used in this research are `R` package in the 2022
PUMF User Guide as unsuitable for trend analysis.

@@

The survey encompasses both rural and urban areas, including
metropolitan and non-metropolitan regions, to ensure a diverse and
representative sample. For sampling, the ten provinces were divided into
geographic strata. Several Census Metropolitan Areas (CMAs) - such as
St. John's, Halifax, Saint John, Montreal, Quebec City, Toronto, Ottawa,
Hamilton, Winnipeg, Regina, Saskatoon, Calgary, Edmonton, and
Vancouver - were treated as separate strata. Additional strata grouped
other CMAs within Quebec, Ontario, and British Columbia, as well as
non-CMA areas within each province.

The Public Use Microdata are released by Statistics Canada in two files:
a *Main file* and an *Episode file*. The files are linked by keys that
identify households, individuals, and episodes (i.e., activities)
conducted by individuals. We discuss these files in more detail in the
following section.

@@

The *Main file* of the time use survey compiles a large array of
aggregated data, summarizing the answers to the questionnaire that
describe households and individuals, as well as derived variables that
summarize the respondents' use of time use across different activities,
locations, and social interactions. This file documents the time and
duration that respondents allocate to each activity and location. The
*Main file* provides a overview of daily routines and social dynamics,
not focusing on individual activity episodes. Additionally, this file
categorizes activities into bigger groups and subcategories,
facilitating the data's analytical utility with additional metrics such
as total transit time, time spent with household members, and counts of
activities and episodes.

Table \ref{tab:main-2015-unprocessed} shows the first ten rows and first
six variables of the TUS PUMF 2015 *Main file* (Cycle 29). Each row in
the table correspond to a survey respondent, while the columns refer the
following information: record identification (`PUMFID`), the person's
weight (`WGHT_PER`), the month the survey data was collected (`SURVMNTH`),
the respondent's age group (`AGEGR10`), the respondent's sex (`SEX`), and
the respondent's marital status (`MARSTAT`).

The *Main file* of the 2015 GSS surveys includes a total of 17,390
respondents, representing `r sum(gss_m_2015$WGHT_PER)` individuals and
848 variables. For discrete variables, Statistics Canada has assigned
specific codes to the possible values, with each code accompanied by a
label. For instance, in the case of the variable `SURVMNTH`, a value of `1`
means `January 2016`, `2` means `February 2016`, `3` corresponds to `March 2016`,
and so on.

@@

```{r gss-main-file-2015}
main_2015_unprocessed <- gss_m_2015[1:10, 1:6]  |>
  kable(
    format = "latex", 
    booktabs = TRUE, 
    longtable = TRUE, 
    align = c(rep("c", ncol(gss_m_2015[1:10, 1:6]))),
    caption = "\\label{tab:main-2015-unprocessed}Visualization of the first ten lines and first six columns of the Main file of the 2015 GSS.") |>
  kable_styling(full_width = FALSE, 
                #latex_options=c("striped", "scale_down"), 
                latex_options=c("scale_down"),
                 font_size = 8) |>
  footnote(general = c("Legend: PUMFID: record identification. WGHT_PER:  person weight. SURVMNTH: survey month of data collection. AGEGR10: age group of the respondent. SEX: sex of the respondent. MARSTAT: marital status of the respondent."), threeparttable = T)

@@

The episode is a much bigger file that records detailed data for each
activity episode reported by respondents. Each episode represents a
single activity and its duration, and the sum of all episodes throughout
the day adds up to 24 hours. Each entry in this file includes the start
and end times of the activity, the duration, location, and accompanying
social context, informing when and where activities occurred and with
whom. The focus of the *Episode file* is not on the characteristics of
the respondents, but on the characteristics of the activities, and the
data are structured around the numerous activity instances that compose
a day of the respondent. Although respondent-specific characteristics
are not included within the *Episode file*, it is possible to link the
*Main file* and the *Episode file* by using a key present in both the
*Main* and *Episode files*.

Similar to Table \ref{tab:main-2015-unprocessed}, which displayed an
example of the *Main file* structure, Table
\ref{tab:ep-2015-unprocessed} presents the first seven episodes for the
record identification number 10041 and some variables from the TUS PUMF
2015 *Episode file* (Cycle 29). Each row in the table corresponds to an
episode associated with the specified record identification (`PUMFID` =
`10041`), including the episode's weight (`WGHT_EPI`), episode number
(`EPINO`), activity code (`TUI_01`), episode duration (`DURATION`), and
episode location (`LOCATION`).

In total, the *Episode file* of the 2015 GSS surveys contains 274,108
records, representing `r sum(gss_e_2015$WGHT_EPI)` episodes and 527
variables detailing the episodes. Similar to the *Main file*, Statistics
Canada has created codes for the discrete variables, with each value
corresponding to a label.

@@

```{r gss-epi-file-2015}
ep_2015_unprocessed <- gss_e_2015 |>
  filter(PUMFID == 10041 & EPINO <= 7) |>
  dplyr::select(PUMFID, EPINO,WGHT_EPI,TUI_01, DURATION, LOCATION) |>
  kable(
    format = "latex", 
    booktabs = TRUE, 
    longtable = TRUE, 
    align = c(rep("c", ncol(gss_m_2015[1:10, 1:6]))),
    caption = "\\label{tab:ep-2015-unprocessed}Visualization of the seven first episodes of the record number 10041.") |>
  kable_styling(full_width = FALSE, 
                #latex_options=c("striped", "scale_down"), 
                latex_options=c("scale_down"),
                 font_size = 8) |>
  footnote(general = c("Legend: PUMFID: record identification. EPINO: episode number. WGHT_EPI: episode's weight. TUI_01: activity code. DURATION: episode's duration. LOCATION: episode's location."), threeparttable = T)

@@

Figure \ref{fig:process-figure} presents a diagram illustrating the
processes applied to the *Main* and *Episode files* to create the
{ActiveCA} datasets. For each TUS cycle, we reviewed the *Episode files*
to identify movement episodes involving walking or cycling. This allowed
us to also select the activities immediately before and after the
movement episode, helping to infer the trip’s purpose and determine its
origin and destination. Active trips were identified by their
corresponding activity codes, accounting for variations across survey
cycles.

@@

For the *Main files*, we selected socioeconomic variables to help
profile individuals engaged in active travel episodes. These included
key indicators such as age group, sex, marital status, and number of
children, among others. As with the *Episode files*, we labeled and
factored the coded variables. Using the appropriate identifiers, we
joined the Episode and *Main files* to determine the province and the
type of population centre - whether it is a CMA, a Census Agglomeration
(CA), or a non-CMA/CA area.

@@

This section presents some potential applications of the {ActiveCA} `R` package. In fact, we expect that the application of this package to extend beyond our pre-imagined range of uses. The installation instruction and also some examples of application of the {ActiveCA} `R` package are available in the `vignettes`, available in the Github repository.

@@

Table \ref{tab:processed-obs} displays the total number of records
processed for Main and *Episode files*. For the *Main files*, a total of
`r sum(df_main_processed$Count)` records were processed, referring to
all records from the Time Use Surveys from 1986 to 2022, that together
represents more of `r sum(df_main_processed$Weighted)` respondents. It
also presents the total cases of active trips episodes identified. In
total `r sum(df_ep_processed$Count)` records with register of active
travel activity. Together, these records account for
`r sum(df_ep_processed$Weighted)` episodes.

@@

table_df_processed <- df_processed_table |> 
  select(-c(Survey_ep, Survey, Year_ep)) %>% 
  kable(format = "latex", booktabs = TRUE, longtable = TRUE, 
        caption = "\\label{tab:processed-obs}Total number and weighted sum of records processed.") |>
  kable_styling(full_width = FALSE, font_size = 8) |>
  add_header_above(c(" " = 1, "Main" = 2, "Episode" = 2 )) %>%
  column_spec(3, border_right = TRUE)

@@

Table \ref{tab:main-2015-processed} shows the first ten rows and first
six variables of the TUS PUMF 2015 Main File (Cycle 29), displayed in
\ref{tab:main-2015-unprocessed} before our processing. Table
\ref{tab:ep-2015-processed} presents the walking episodes for the record
identification number `10041` from the TUS PUMF 2015 *Episode file* (Cycle
29), previously displayed in Table \ref{tab:ep-2015-unprocessed}. Only
the unique active travel episode appear in Table
\ref{tab:ep-2015-processed} since the records were filtered to select
cases with walking or cycling episodes. For both cases, Tables
\ref{tab:main-2015-processed} and \ref{tab:ep-2015-processed} contain
labeled variables, facilitating the interpretation of the data.

```{r gss-processed-file-2015, echo=FALSE, cache=FALSE}
main_2015_processed <- gss_main_2015[1:10, 1:6]  |>
  kable(
    format = "latex", 
    booktabs = TRUE, 
    longtable = TRUE, 
    align = c(rep("c", ncol(gss_m_2015[1:10, 1:6]))),
    caption = "\\label{tab:main-2015-processed}Visualization of the first ten lines and first six columns of the 2015 TUS Main File.") |>
  kable_styling(full_width = FALSE, 
                #latex_options=c("striped", "scale_down"), 
                latex_options=c("scale_down"),
                 font_size = 8) |>
  footnote(general = c("Legend: PUMFID: record identification. WGHT_PER:  person weight. SURVMNTH: survey month of data collection. AGEGR10: age group of the respondent. SEX: sex of the respondent. MARSTAT: marital status of the respondent."), threeparttable = T)

@@

ep_2015_unprocessed <- gss_episodes_renamed |>
  filter(PUMFID == 10041 & YEAR == 2015) |>
  dplyr::select(PUMFID, WGHT_EPI, act_label, DURATION, orig_label, dest_label, MODE) |>
  kable(
    format = "latex", 
    booktabs = TRUE, 
    longtable = TRUE, 
    caption = "\\label{tab:ep-2015-processed}Visualization of the active travel episode for the record number 10041 of the 2015 GSS survey.", 
       col.names = c("PUMFID", "WGHT_EPI","Activity","Duration","Origin", "Destination", "Mode")) |> 
  kable_styling(full_width = FALSE, 
                #latex_options=c("striped", "scale_down"), 
                latex_options=c("scale_down"),
                 font_size = 8) |>
  footnote(general = c("Legend: PUMFID: record identification. EPINO: episode number. WGHT_EPI: episode's weight. TUI_01: activity code. DURATION: episode's duration. LOCATION: episode's location."), threeparttable = T)

@@

# Creating a latex table
styled_ch03_table_01 <- stats_data |>
  kable(format = "latex", booktabs = TRUE, longtable = TRUE, 
        caption = "\\label{tab:table-01}Descriptive statistics for episodes with active transport records",  
        align = c('l', 'l', rep('c', ncol(stats_data) - 2))) |>
  kable_styling(full_width = FALSE, font_size = 8) |>
  column_spec(1, bold = TRUE) |>
  add_header_above(c(" " = 2, "Year" = ncol(stats_data) - 2)) |>
  collapse_rows(columns = 1, valign = "top", latex_hline = "major")

@@

In 2022, `home` served as a central hub for most trips, with fewer than
10% of journeys not involving it as either a starting point or
destination. The most common trip types were from `home` to `work or school`
(17%) and the reverse, from `work or school` to `home` (13%). Notably, 7% of
trips began and ended at `home`, often reflecting leisure activities such
as short walks or dog walking. `Grocery stores` were also a key
destination, comprising 10% of trips departing from `home.`

@@

# Latex table
median_duration_table <- median_duration_cma |>
  kable(
    format = "latex", 
    booktabs = TRUE, 
    caption = "\\label{tab:cma-durations}Differences in walking duration between provinces and population centre type.") |>
  kable_styling(full_width = FALSE, 
                #latex_options=c("striped", "scale_down"), 
                latex_options=c("scale_down"),
                 font_size = 8) |>
  add_header_above(c(" " = 1, "Population centre type" = 2)) |> 
  footnote(general = c("CMA denotes Census Metropolitan Area and CA denotes Census agglomeration."), threeparttable = T)

@@

{ActiveCA} also provides a Jupyter Notebook containing a `Python` script
that demonstrates how to read `R` data files (.rda) and convert them into
Pandas DataFrames. This process allows users to work with and utilize
the datasets available in {ActiveCA} within a `Python` project.

@@

This paper presents {ActiveCA}, an open data product that provides
analysis-ready data from Cycles 2 (1986), 7 (1992), 12 (1998), 19
(2005), 24 (2010), 29 (2015), and 34 (2022) of TUS GSSs on active travel
in Canada. In the form of an `R` data package, {ActiveCA} was developed
after collecting, cleaning, and processing the survey data, providing
information on origins, destinations, and duration of active travel, as
well other information.

@@

The author(s) declared no potential conflicts of interest with respect
to the research, authorship, and/or publication of this article.

@@

The author(s) disclosed receipt of the following financial support for
the research, authorship, and/or publication of this article: This work
was supported by the Social Sciences and Humanities Research Council of
Canada (*More description about the funding source after the review
process*).

@@

Author 1

Author 2

Author 3

@@

The {ActiveCA} R data package can be found and installed on Github
(*link*).

For review purposes, the package is currently available as a tar.gz file
that can be installed by R. The file can be obtained from this anonymous
location:

<https://user.fm/files/v2-3d261d0b2aa47fcad096cd9e49fd5cf8/ActiveCA.zip>

>>>>

---
title: "ActiveCA: Time Use Data from the General Social Survey of Canada to Study Active Travel"
runninghead: Dos Santos \emph{et al}.
author:
- name: Bruno Dias dos Santos*
  num: 1
- name: Mahdis Moghadasi
  num: 1 
- name: Antonio Páez
  num: 1
address:
- num: 1
  org: McMaster University
corrauth: "Bruno Dias dos Santos - School of Earth, Environment and Society, McMaster University, 1280 Main St W, Hamilton, Ontario L8S 4K1 Canada."
email: dossanb@mcmaster.ca
abstract: 
  "This paper describes {ActiveCA}, an open data product with Canadian time use data. {ActiveCA} is an R data package that contains analysis-ready data related to active travel spanning almost 40 years, extracted from Cycles 2 (1986), 7 (1992), 12 (1998), 19 (2005), 24 (2010), 29 (2015), and 34 (2022) of the Time Use Survey (TUS) from the General Social Survey (GSS). Active travel episodes are characterized by mode, with walking being part of every cycle and bicycling starting in 1992. The attributes of active trips are the types of locations of origins and destinations, the duration of trips, and episode weights for expanding the trips to population-wide estimates. Based on the year of the survey, a variety of locations are coded. In earlier cycles, these include home, work or school, and other's home, whereas in later cycles these are augmented with locations such as grocery stores, restaurants, outdoor destinations, and others. The geographical resolution includes the province and whether the episode was in an urban or rural setting." 
link-citations: yes
linkcolor: blue
keywords: Active; mobility; walking; cycling; travel time; time-use; 
classoption:
  - Royal
  - times
bibliography: bibfile.bib
bibliographystyle: sageh
output:
  rticles::sage_article:
    keep_tex: yes
editor_options: 
  markdown: 
    wrap: 72
---

@@

{ActiveCA} provides analysis-ready data concerning active travel in
Canada spanning a period of almost 40 years, obtained from the TUS
cycles of the Canadian GSS. The GSS program is designed to provide
cross-sectional data on topics of interest to improve the well-being of
Canadians. As part of this program, every five to seven years the survey
is done on the topic of time use. Concretely, {ActiveCA} covers
Cycles 2 (1986), 7 (1992), 12 (1998), 19 (2005), 24 (2010), 29 (2015),
and 34 (2022) of the TUS. Time use data in these surveys is coded using
a very fine grain, from time spent in chores, leisure, and sleeping, to
time spent working or at school. These surveys have proved valuable in
investigations of mobility and quality of life [@spinneyTransport2009],
the relationship between active travel and transit use
[@lachapelleLonger2016], and travel behavior and time poverty
[@kimFacing2024], to name but a few examples.

Using Public Use Microdata Files (PUMFs) from the TUS,
we extracted all data necessary to characterize active travel in
Canada - specifically, episodes in which the activity involved moving
between an origin and a destination by walking or cycling. Although
Statistics Canada provides PUMFs and accompanying documentation for the
GSS program [see @statisticscanada2024], accessing and preparing these
files for analysis is not a straightforward task due to their size and
complexity. The process of extracting information of interest from the
source files is time-consuming, tedious, and challenging and/or prone to
error due to the expertise required to work with these files. To create
{ActiveCA} we selected, labelled, and processed the TUS cycles to make
them ready for analysis.

{ActiveCA} is distributed as an R package with a number of data objects
and their documentation. R packages contain code, data, and
documentation in a standardized format that can be installed by R users
via a software repository, such as CRAN (Comprehensive R Archive
Network) or GitHub, which makes them an adroit medium to distribute
analysis-ready data.

@@

The rest of this paper discusses the sources of data, and the process
implemented to retrieve and package them. Then, we show some selected examples of analysis to whet the
imagination of potential users. This ODP provides not only data that are
easy to use, but also all the code and documentation that make this a
reproducible research project. In summary, {ActiveCA} aims to implement
and inspire the best principles of open spatial sciences
[@paez_open_2021; @brunsdon_opening_2021].

@@

Most respondents to the 2022 TUS completed it online, reflecting
Statistics Canada's effort to adapt to technological changes and growing
time demands by offering greater flexibility and convenience
[@statisticscanada2022]. While such methodological changes may affect
data comparability over time, it is not possible to determine whether
observed differences result from actual population changes or shifts in
data collection methods. Despite rigorous efforts to ensure data
quality, the use of electronic questionnaires may have influenced
estimates. Statistics Canada assessed the impact of collection mode on a
limited set of key questions, constrained by sample size. Importantly,
none of the variables used in this research are in the 2022
PUMF User Guide as unsuitable for trend analysis.

@@

The survey encompasses both rural and urban areas, including
metropolitan and non-metropolitan regions, to ensure a diverse and
representative sample. For sampling, the ten provinces were divided into
geographic stratas. Several Census Metropolitan Areas (CMAs) - such as
St. John's, Halifax, Saint John, Montreal, Quebec City, Toronto, Ottawa,
Hamilton, Winnipeg, Regina, Saskatoon, Calgary, Edmonton, and
Vancouver - were treated as separate strata. Additional strata grouped
other CMAs within Quebec, Ontario, and British Columbia, as well as
non-CMA areas within each province.

The Public Use Microdata are released by Statistics Canada in two files:
a main file and an episode file. The files are linked by keys that
identify households, individuals, and episodes (i.e., activities)
conducted by individuals. We discuss these files in more detail in the
following section.

@@

The main file of the time use survey compiles a large array of
aggregated data, summarizing the answers to the questionnaire that
describe households and individuals, as well as derived variables that
summarize the respondents' use of time use across different activities,
locations, and social interactions. This file documents the time and
duration that respondents allocate to each activity and location. The
main file provides a overview of daily routines and social dynamics,
not focusing on individual activity episodes. Additionally, this file
categorizes activities into bigger groups and subcategories,
facilitating the data's analytical utility with additional metrics such
as total transit time, time spent with household members, and counts of
activities and episodes.

Table \ref{tab:main-2015-unprocessed} shows the first ten rows and first
six variables of the TUS PUMF 2015 main file (Cycle 29). Each row in
the table correspond to a survey respondent, while the columns refer the
following information: record identification (`PUMFID`), the person's
weight (`WGHT_PER`), the month the survey data was collected (`SURVMNTH`),
the respondent's age group (`AGEGR10`), the respondent's sex (`SEX`), and
the respondent's marital status (`MARSTAT`).

The main file of the 2015 GSS surveys includes a total of 17,390
respondents, representing `r sum(gss_m_2015$WGHT_PER)` individuals and
848 variables. For discrete variables, Statistics Canada has assigned
specific codes to the possible values, with each code accompanied by a
label. For instance, in the case of the variable `SURVMNTH`, a value of `1`
means `January 2016`, `2` means `February 2016`, `3` corresponds to `March 2016`,
and so on.

@@

```{r gss-main-file-2015}
main_2015_unprocessed <- gss_m_2015[1:10, 1:6]  |>
  kable(
    format = "latex", 
    booktabs = TRUE, 
    longtable = FALSE, 
    align = c(rep("c", ncol(gss_m_2015[1:10, 1:6]))),
    caption = "\\label{tab:main-2015-unprocessed}Visualization of the first ten lines and first six columns of the Main file of the 2015 GSS.") |>
  kable_styling(full_width = FALSE, 
                #latex_options=c("striped", "scale_down"), 
                latex_options=c("scale_down"),
                 font_size = 8) |>
  footnote(general = c("Legend: PUMFID: record identification. WGHT_PER: person weight. SURVMNTH: survey month of data collection. AGEGR10: age group of the respondent. SEX: sex of the respondent. MARSTAT: marital status of the respondent."), threeparttable = T)

@@

The episode is a much bigger file that records detailed data for each
activity episode reported by respondents. Each episode represents a
single activity and its duration, and the sum of all episodes throughout
the day adds up to 24 hours. Each entry in this file includes the start
and end times of the activity, the duration, location, and accompanying
social context, informing when and where activities occurred and with
whom. The focus of the episode file is not on the characteristics of
the respondents, but on the characteristics of the activities, and the
data are structured around the numerous activity instances that compose
a day of the respondent. Although respondent-specific characteristics
are not included within the episode file, it is possible to link the
main file and the episode file by using a key present in both the
main and episodes files.

Similar to Table \ref{tab:main-2015-unprocessed}, which displayed an
example of the main file structure, Table
\ref{tab:ep-2015-unprocessed} presents the first seven episodes for the
record identification number 10041 and some variables from the TUS PUMF
2015 episode file (Cycle 29). Each row in the table corresponds to an
episode associated with the specified record identification (`PUMFID` =
`10041`), including the episode's weight (`WGHT_EPI`), episode number
(`EPINO`), activity code (`TUI_01`), episode duration (`DURATION`), and
episode location (`LOCATION`).

In total, the episode file of the 2015 GSS surveys contains 274,108
records, representing `r sum(gss_e_2015$WGHT_EPI)` episodes and 527
variables detailing the episodes. Similar to the main file, Statistics
Canada has created codes for the discrete variables, with each value
corresponding to a label.

@@

```{r gss-epi-file-2015}
ep_2015_unprocessed <- gss_e_2015 |>
  filter(PUMFID == 10041 & EPINO <= 7) |>
  dplyr::select(PUMFID, EPINO,WGHT_EPI,TUI_01, DURATION, LOCATION) |>
  kable(
    format = "latex", 
    booktabs = TRUE, 
    longtable = FALSE, 
    align = c(rep("c", ncol(gss_m_2015[1:10, 1:6]))),
    caption = "\\label{tab:ep-2015-unprocessed}Visualization of the  first seven episodes of the record number 10041.") |>
  kable_styling(full_width = FALSE, 
                #latex_options=c("striped", "scale_down"), 
                latex_options=c("scale_down"),
                 font_size = 8) |>
  footnote(general = c("Legend: PUMFID: record identification. EPINO: episode number. WGHT_EPI: episode's weight. TUI_01: activity code. DURATION: episode's duration. LOCATION: episode's location."), threeparttable = T)

@@

Figure \ref{fig:process-figure} presents a diagram illustrating the
processes applied to the main and episode files to create the
{ActiveCA} datasets. For each TUS cycle, we reviewed the episode files
to identify movement episodes involving walking or cycling. This allowed
us to also select the activities immediately before and after the
movement episode, helping to infer the trip’s purpose and determine its
origin and destination. Active trips were identified by their
corresponding activity codes, accounting for variations across survey
cycles.

@@

For the main files, we selected socioeconomic variables to help
profile individuals engaged in active travel episodes. These included
key indicators such as age group, sex, marital status, and number of
children, among others. As with the episode files, we labeled and
factored the coded variables. Using the appropriate identifiers, we
joined the episode and main files to determine the province and the
type of population centre - whether it is a CMA, a Census Agglomeration
(CA), or a non-CMA/CA area.

@@

This section presents some potential applications of the {ActiveCA} R package. In fact, we expect that the application of this package to extend beyond our pre-imagined range of uses. The installation instruction and also some examples of application of the {ActiveCA} R package are available in the vignettes, available in the Github repository.

@@

Table \ref{tab:processed-obs} displays the total number of records
processed for Main and episode files. For the main files, a total of
`r sum(df_main_processed$Count)` records were processed, referring to
all records from the Time Use Surveys from 1986 to 2022, that together
represents more of `r sum(df_main_processed$Weighted)` respondents. It
also presents the total cases of active trips episodes identified. In
total `r sum(df_ep_processed$Count)` records with register of active
travel activity. Together, these records account for
`r sum(df_ep_processed$Weighted)` episodes.

@@

table_df_processed <- df_processed_table |> 
  select(-c(Survey_ep, Survey, Year_ep)) %>% 
  kable(format = "latex", booktabs = TRUE, longtable = FALSE, 
        caption = "\\label{tab:processed-obs}Total number and weighted sum of records processed.") |>
  kable_styling(full_width = FALSE, font_size = 8) |>
  add_header_above(c(" " = 1, "Main" = 2, "Episode" = 2 )) %>%
  column_spec(3, border_right = TRUE)

@@

Table \ref{tab:main-2015-processed} shows the first ten rows and first
six variables of the TUS PUMF 2015 Main File (Cycle 29), displayed in
\ref{tab:main-2015-unprocessed} before our processing. Table
\ref{tab:ep-2015-processed} presents the walking episodes for the record
identification number `10041` from the TUS PUMF 2015 episode file (Cycle
29), previously displayed in Table \ref{tab:ep-2015-unprocessed}. Only
the unique active travel episode appears in Table
\ref{tab:ep-2015-processed} since the records were filtered to select
cases with walking or cycling episodes. For both cases, Tables
\ref{tab:main-2015-processed} and \ref{tab:ep-2015-processed} contain
labeled variables, facilitating the interpretation of the data.

```{r gss-processed-file-2015, echo=FALSE, cache=FALSE}
main_2015_processed <- gss_main_2015[1:10, 1:6]  |>
  kable(
    format = "latex", 
    booktabs = TRUE, 
    longtable = FALSE, 
    align = c(rep("c", ncol(gss_m_2015[1:10, 1:6]))),
    caption = "\\label{tab:main-2015-processed}Visualization of the first ten lines and first six columns of the 2015 TUS Main File.") |>
  kable_styling(full_width = FALSE, 
                #latex_options=c("striped", "scale_down"), 
                latex_options=c("scale_down"),
                 font_size = 8) |>
  footnote(general = c("Legend: PUMFID: record identification. WGHT_PER:  person weight. SURVMNTH: survey month of data collection. AGEGR10: age group of the respondent. SEX: sex of the respondent. MARSTAT: marital status of the respondent."), threeparttable = T)

@@

ep_2015_unprocessed <- gss_episodes_renamed |>
  filter(PUMFID == 10041 & YEAR == 2015) |>
  dplyr::select(PUMFID, WGHT_EPI, act_label, DURATION, orig_label, dest_label, MODE) |>
  kable(
    format = "latex", 
    booktabs = TRUE, 
    longtable = FALSE, 
    caption = "\\label{tab:ep-2015-processed}Visualization of the active travel episode for the record number 10041 of the 2015 GSS survey.", 
       col.names = c("PUMFID", "WGHT_EPI","Activity","Duration","Origin", "Destination", "Mode")) |> 
  kable_styling(full_width = FALSE, 
                #latex_options=c("striped", "scale_down"), 
                latex_options=c("scale_down"),
                 font_size = 8) |>
  footnote(general = c("Legend: PUMFID: record identification. EPINO: episode number. WGHT_EPI: episode's weight. TUI_01: activity code. DURATION: episode's duration. LOCATION: episode's location."), threeparttable = T)

@@

# Creating a latex table
styled_ch03_table_01 <- stats_data |>
  kable(format = "latex", booktabs = TRUE, longtable = FALSE, 
        caption = "\\label{tab:table-01}Descriptive statistics of the duration (in minutes) of episodes with active transport records.",  
        align = c('l', 'l', rep('c', ncol(stats_data) - 2))) |>
  kable_styling(full_width = FALSE, font_size = 8) |>
  column_spec(1, bold = TRUE) |>
  add_header_above(c(" " = 2, "Year" = ncol(stats_data) - 2)) |>
  collapse_rows(columns = 1, valign = "top", latex_hline = "major")

@@

In 2022, home location served as a central hub for most trips, with fewer than
10% of journeys not involving it as either a starting point or
destination. The most common trip types were from home to work or school
(17%) and the reverse, from work or school to home (13%). Notably, 7% of
trips began and ended at home, often reflecting leisure activities such
as short walks or dog walking. Grocery stores were also a key
destination, comprising 10% of trips departing from home.

@@

# Latex table
median_duration_table <- median_duration_cma |>
  kable(
    format = "latex", 
    booktabs = TRUE, 
    caption = "\\label{tab:cma-durations}Differences in walking duration (in minutes) between provinces and population centre type.") |>
  kable_styling(full_width = FALSE, 
                #latex_options=c("striped", "scale_down"), 
                latex_options=c("scale_down"),
                 font_size = 8) |>
  add_header_above(c(" " = 1, "Population centre type" = 2)) |> 
  footnote(general = c("CMA denotes Census Metropolitan Area and CA denotes Census agglomeration."), threeparttable = T)

@@

{ActiveCA} also provides a Jupyter Notebook containing a Python script
that demonstrates how to read R data files (.rda) and convert them into
Pandas Data Frames. This process allows users to work with and utilize
the datasets available in {ActiveCA} within a Python project.

@@

This paper presents {ActiveCA}, an open data product that provides
analysis-ready data from Cycles 2 (1986), 7 (1992), 12 (1998), 19
(2005), 24 (2010), 29 (2015), and 34 (2022) of TUS GSSs on active travel
in Canada. In the form of an R data package, {ActiveCA} was developed
after collecting, cleaning, and processing the survey data, providing
information on origins, destinations, and duration of active travel, as
well other information.

@@

The authors declared no potential conflicts of interest with respect
to the research, authorship, and/or publication of this article.

# Acknowledgments

The authors thank the anonymous reviewers for their insightful comments.

@@

The work has been supported by funding from the Social Sciences and Humanities Research Council’s Partnership Grant: Mobilizing justice: towards evidence-based transportation equity policy.

@@

Bruno Dias dos Santos - https://orcid.org/0000-0002-6748-2038

Mahdis Moghadasi - https://orcid.org/0000-0003-1009-9369

Antonio Páez - https://orcid.org/0000-0001-6912-9919

@@

The {ActiveCA} R data package can be found and installed on [Github](https://github.com/dias-bruno/ActiveCA).
